---
title: "Batch Geocoding with Census API"
output:
  html_document:
    theme: readable
    df_print: paged
    highlight: tango
    css: textbook.css
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=F, message=F, fig.width=8)
```


### packages

```{r}
library( dplyr )
library( tidyr )
library( httr )

```



In this step we will geocode addresses of the founding teams (managers and board members) and the nonprofit addresses.

***Structure***

1. Intro and Demo
PEOPLE
2. Geocoding
3. Compiling addresses
NPOs
4. Geocoding
5. Compiling addresses




# Intro to the Batch Geocoder and Demo


**CENSUS GEOCODER** 

DOCUMENTATION:
https://www.census.gov/programs-surveys/geography.html

WEB FORM:
https://geocoding.geo.census.gov/geocoder/geographies/addressbatch?form

Addresses should be formatted in a single line with comma delimiters. The address should consist of:

* Unique ID,
* House Number and Street Name,
* City,
* State,
* ZIP Code


```{r}
dat.people <- readRDS( "Data/PEOPLE-2014-2018.rds" )
head( dat.people, 20 )

#selecting the addresses and placing them in a new df.
addr <- select( dat.people, ID, Ofcrdirtruststreetaddr, Ofcrdirtrustcity, Ofcrdirtruststate, Ofcrdirtrustzip )
names( addr ) <- c("ID", "STREET", "CITY", "STATE", "ZIP" )

#test df
addr.test <- addr[ 1:20, ]
write.csv( addr.test, "TestAddresses.csv", row.names=F )



### RETURN GEOGRAPHIES (lat/lon plus census tracts & blocks)
### library( httr )

apiurl <- "https://geocoding.geo.census.gov/geocoder/geographies/addressbatch"

addressFile <- "TestAddresses.csv"

resp <- POST( apiurl, 
              body=list(addressFile=upload_file(addressFile), 
                        benchmark="Public_AR_Census2010",
                        vintage="Census2010_Census2010",
                        returntype="csv" ), 
              encode="multipart" )

# content( resp, as="text" )
# content( resp )

results <- "results.csv"
var_names <- c( "id", "input_address", 
                "match", "match_type", 
                "out_address", "lat_lon", 
                "tiger_line_id", "tiger_line_side", 
                "state_fips", "county_fips", 
                "tract_fips", "block_fips" )
v.names <- paste(var_names, collapse=',')
writeLines( text=c(v.names, content(resp)), con="results.csv" )



### ADD VARIABLE NAMES
# from: https://www.census.gov/geo/maps-data/data/geocoder.html

res <- read.csv( "results.csv", header=T, stringsAsFactors=F, colClasses="character" )


### SPLIT LATITUDE AND LONGITUDE COORDINATES

lat.lon <- strsplit( res$lat_lon, "," )

for( i in 1:length(lat.lon) )
{
   # print( length( lat.lon[[i]] ) )
   if( length( lat.lon[[i]] ) < 2 )
   lat.lon[[ i ]] <- c(NA,NA) 
}

m <- matrix( unlist( lat.lon ), ncol=2, byrow=T )
colnames(m) <- c("lon","lat")
m <- as.data.frame( m )

res <- cbind( res, m )
head( res )

# write.csv( res, "ResultsAugmented.csv", row.names=F )
```



# 2. BATCH GEOCODE NONPROFITS

copying code for people, adapt for nonprofit addresses

```{r, eval=F}
setwd("/Users/icps86/Documents/R Projects/Open Data Project/open-1023-ez-dataset-master")
getwd()
dat.npos <- readRDS( "Data/NONPROFIT-ADDRESSES-2014-2018.rds" )
```



## 2.1: getting the nonprofit data in batches files.

```{r}
d2 <- dat.npos #this is a backup

# To Geocode we need to format the DF into one that has the following input file variables: 
# ID	STREET	CITY	STATE	ZIP

#str(dat.npos)
dat.npos <- select( dat.npos, ID, Address, City, State, Zip )

#now we create a folder and set the working dir inside that folder.
dir.create( "addresses_npos" )
setwd( "addresses_npos" )
#getwd()
#dir()

names( dat.npos ) <- NULL  # we remove colnames because input file should not have names

# split address files into files with 500 addresses each
loops <- ceiling( nrow( dat.npos ) / 500 ) # ceiling function rounds up an integer. so loops has the amount of 500s that fit rounded up.

# SPLIT DATA INTO PARTS
for( i in 1:loops ) #now we use a loop to extract by 500 addresses.
{
filename <- paste0( "AddressNpos",i,".csv" )
start.row <- ((i-1)*500+1) # i starts in 1 and this outputs: 1, 501, 1001, etc.
end.row <- (500*i) #this outputs 500, 1000, etc.
if( nrow(dat.npos) < end.row ){ end.row <- nrow(dat.npos) } #this is the "fusible" that makes the end.row variable be set to the actual end row in the last iteration of the loop.

#now we write the piece of the address file.
write.csv( dat.npos[ start.row:end.row, ], filename, row.names=F )

#what follows is just an output to help keep track of the loop.
print( i )
print( paste( "Start Row:", start.row ) )
print( paste( "End Row:", end.row ) )
} #end loop.

```


## 2.2 START GEOCODING

The following code will produce 2 batches of files in addition to the AddressPeopleXX.csv that already exists in the folder.
One is ResultsXX, and the other is ResultspeopleXX, plus, a text and csv for the log.

Before running the code, make sure the working dir is correct and the loop variable is in the environment. In this iteration the value is 1511.

```{r}
### RETURN GEOGRAPHIES (lat/lon plus census tracts & blocks)

# setwd('..')
# getwd()
#producing LOG file
log <- c("Query_Number", "Start_time", "Time_taken")
log <- paste(log, collapse=',')
write(log, file="Geocode_Log.txt", append = F)


#make sure the loop starts where you want it to.
dir()
for( i in 1:loops )
{ # LOOP STARTS

#creating the objects needed to run the loop.
  
filename <- paste0( "AddressNpos",i,".csv" ) # constructing name of file to load

addressFile <- filename

apiurl <- "https://geocoding.geo.census.gov/geocoder/geographies/addressbatch"

#starting first goecoding, first we output the time, also at the end of this step.
print( i )
print(Sys.time() )
start_time <- Sys.time()

try( #try is a wrapper to run an expression that might fail and allow the users code to handle error-recovery.
  
resp <- POST( apiurl, 
              body=list(addressFile=upload_file(addressFile), 
                        benchmark="Public_AR_Census2010",
                        vintage="Census2010_Census2010",
                        returntype="csv" ), 
              encode="multipart" )
  )

#printing ending times for monitoring loop
end_time <- Sys.time()
print( end_time - start_time ) #ouputting in R console

#writing in log.
query <- as.character(i)
len <- as.character(end_time - start_time)
start_time <- as.character(start_time)
log <- c(query, start_time, len)
log <- paste(log, collapse=',')
write(log, file="Geocode_Log.txt", append = T)

### ADD VARIABLE NAMES
# from: https://www.census.gov/geo/maps-data/data/geocoder.html

filename2 <- paste0( "results",i,".csv" ) #constructing name of file to write.

results <- filename2

var_names <- c( "id", "input_address", 
                "match", "match_type", 
                "out_address", "lat_lon", 
                "tiger_line_id", "tiger_line_side", 
                "state_fips", "county_fips", 
                "tract_fips", "block_fips" )
v.names <- paste(var_names, collapse=',')
writeLines( text=c(v.names, content(resp)) , con=filename2 )# writting the first output file as filename2 which is named as resultsXX.csv

# writeLines( text=content( resp ) , con=filename2 )  

res <- read.csv( filename2, header=T, 
                 stringsAsFactors=F, 
                 colClasses="character" )


### SPLIT LATITUDE AND LONGITUDE COORDINATES

lat.lon <- strsplit( res$lat_lon, "," )

for( j in 1:length(lat.lon) )
{
   # print( length( lat.lon[[j]] ) )
   if( length( lat.lon[[j]] ) < 2 )
   lat.lon[[ j ]] <- c(NA,NA) 
}

m <- matrix( unlist( lat.lon ), ncol=2, byrow=T )
colnames(m) <- c("lon","lat")
m <- as.data.frame( m )

res <- cbind( res, m )

write.csv( res, paste0("ResultsNpos",i,".csv"), row.names=F )

}  # end of loop

getwd()
```

1st NON PROFIT ADDRESS GEOCODE.
might have stopped at 9. because of computer locking itself.
might have stopped at 303

2nd GEOCODE
start 7.55pm.

12.14pm 366 en el loop.
419 ended at 1:51


## 2.3 Combining all the Files

```{r}
setwd("/Users/icps86/Documents/R Projects/Open Data Project/open-1023-ez-dataset-master")
setwd("addresses_npos")
setwd("Results4")
getwd()
dir()
these <- (dir())[ grepl( "ResultsPeople", dir() ) ] #this gets the names of all the files that have the string ResultsPeople
#I think the expression (dir()) makes the output a list

dat <- read.csv( these[1], stringsAsFactors=F ) #this generates an object with the first observation.


for( i in 2:length(these) )
{

  d <- read.csv( these[i], stringsAsFactors=F )
  dat <- bind_rows( dat, d ) 

}

saveRDS( dat, "../../GeocodeResults/NPOs-GEOCODESv4.rds" )

plot( dat$lon, dat$lat, pch=19, cex=0.5, col=gray(0.5,0.01) )

```

* PEOPLE-2014-2018.rds
* NONPROFIT-ADDRESSES-2014-2018.rds

Through the Census Service

* PEOPLE-GEOCODESv1.rds
* PEOPLE-GEOCODESv2.rds

* NPOs-GEOCODESv1.rds
* NPOs-GEOCODESv2.rds
* NPOs-GEOCODESv3.rds
* NPOs-GEOCODESv4.rds

Compiled (rmd script compiling geocodes) the results into two files:

* NPOS-GEOCODES-COMPILED.rds

* PEOPLE-GEOCODES-COMPILED.rds



# 3. BATCH GEOCODE PEOPLE

```{r, eval=F}
#we are loading the DF with non profit People data
dat.people <- readRDS( "Data/PEOPLE-2014-2018.rds" )
```


"Batch files may not exceed 10,000 records."

## 3.1: getting the people data in batches files.

```{r}
d2 <- dat.people

# To Geocode we need to format the DF into one that has the following input file variables: 
# ID	STREET	CITY	STATE	ZIP

dat.people <- select( dat.people, ID, Ofcrdirtruststreetaddr, 
              Ofcrdirtrustcity, Ofcrdirtruststate, Ofcrdirtrustzip )

#now we create a folder and set the working dir inside that folder.
dir.create( "addresses_people" )
setwd( "addresses_people" )
#getwd()
#dir()

names( dat.people ) <- NULL  # we remove colnames because input file should not have names

# split address files into files with 500 addresses each
loops <- ceiling( nrow( dat.people ) / 500 ) # ceiling function rounds up an integer. so loops has the amount of 500s that fit rounded up.

# SPLIT DATA INTO PARTS
for( i in 1:loops ) #now we use a loop to extract by 500 addresses.
{
filename <- paste0( "AddressesPeople",i,".csv" )
start.row <- ((i-1)*500+1) # i starts in 1 and this outputs: 1, 501, 1001, etc.
end.row <- (500*i) #this outputs 500, 1000, etc.
if( nrow(dat.people) < end.row ){ end.row <- nrow(dat.people) } #this is the "fusible" that makes the end.row variable be set to the actual end row in the last iteration of the loop.

#now we write the piece of the address file.
write.csv( dat.people[ start.row:end.row, ], filename, row.names=F )

#what follows is just an output to help keep track of the loop.
print( i )
print( paste( "Start Row:", start.row ) )
print( paste( "End Row:", end.row ) )
} #end loop.

```


## 3.2 START GEOCODING

The following code will produce 2 batches of files in addition to the AddressPeopleXX.csv that already exists in the folder.
One is ResultsXX, and the other is ResultspeopleXX, plus, a text and csv for the log.

Before running the code, make sure the working dir is correct and the loop variable is in the environment. In this iteration the value is 1511.

```{r}
### RETURN GEOGRAPHIES (lat/lon plus census tracts & blocks)


#setwd("/Users/icps86/Documents/R Projects/Open Data Project/open-1023-ez-dataset-master")
#dat.people <- readRDS( "Data/PEOPLE-2014-2018.rds" )
#dat.people <- select( dat.people, ID, Ofcrdirtruststreetaddr, Ofcrdirtrustcity, Ofcrdirtruststate, Ofcrdirtrustzip )
#loops <- ceiling( nrow( dat.people ) / 500 )
#for test: loops <- 2

# NOTE: for this to work it needs to have 
# setwd( "addresses_people" ) 
# setwd('..')
# getwd()

#producing LOG file
log <- c("Query_Number", "Start_time", "Time_taken")
log <- paste(log, collapse=',')
write(log, file="Geocode_Log.txt", append = F)


#make sure the loop starts where you want it to.

for( i in 1:loops )
{ # LOOP STARTS

#creating the objects needed to run the loop.
  
filename <- paste0( "AddressesPeople",i,".csv" ) # constructing name of file to load

addressFile <- filename

apiurl <- "https://geocoding.geo.census.gov/geocoder/geographies/addressbatch"

#starting first goecoding, first we output the time, also at the end of this step.
print( i )
print(Sys.time() )
start_time <- Sys.time()

try( #try is a wrapper to run an expression that might fail and allow the users code to handle error-recovery.
  
resp <- POST( apiurl, 
              body=list(addressFile=upload_file(addressFile), 
                        benchmark="Public_AR_Census2010",
                        vintage="Census2010_Census2010",
                        returntype="csv" ), 
              encode="multipart" )
)

#printing ending times for monitoring loop
end_time <- Sys.time()
print( end_time - start_time ) #ouputting in R console

#writing in log.
query <- as.character(i)
len <- as.character(end_time - start_time)
start_time <- as.character(start_time)
log <- c(query, start_time, len)
log <- paste(log, collapse=',')
write(log, file="Geocode_Log.txt", append = T)

################

### ADD VARIABLE NAMES
# from: https://www.census.gov/geo/maps-data/data/geocoder.html

filename2 <- paste0( "results",i,".csv" ) #constructing name of file to write.

results <- filename2

var_names <- c( "id", "input_address", 
                "match", "match_type", 
                "out_address", "lat_lon", 
                "tiger_line_id", "tiger_line_side", 
                "state_fips", "county_fips", 
                "tract_fips", "block_fips" )
v.names <- paste(var_names, collapse=',')
writeLines( text=c(v.names, content(resp)) , con=filename2 )# writting the first output file as filename2 which is named as resultsXX.csv

# writeLines( text=content( resp ) , con=filename2 )  

res <- read.csv( filename2, header=T, 
                 stringsAsFactors=F, 
                 colClasses="character" )


### SPLIT LATITUDE AND LONGITUDE COORDINATES

lat.lon <- strsplit( res$lat_lon, "," )

for( j in 1:length(lat.lon) )
{
   # print( length( lat.lon[[j]] ) )
   if( length( lat.lon[[j]] ) < 2 )
   lat.lon[[ j ]] <- c(NA,NA) 
}

m <- matrix( unlist( lat.lon ), ncol=2, byrow=T )
colnames(m) <- c("lon","lat")
m <- as.data.frame( m )

res <- cbind( res, m )

write.csv( res, paste0("ResultsPeople",i,".csv"), row.names=F )

}  # end of loop

getwd()
```

Two days of geocoding:

#Tuesday
Run at TUE 11.43pm
Stopped at WED 8.26am
finished 539 of the loop

#Wednesday
Start at 539 (repiting this one for later check in consistency)
in the morning and had finished by the next day.


## 3.3 Combining all the Files

```{r}
setwd("../addresses_people/Results")
setwd("..")
setwd("../addresses_npos")

getwd()
dir()

these <- (dir())[ grepl( "ResultsPeople", dir() ) ] #this gets the names of all the files that have the string ResultsPeople
#I think the expression (dir()) makes the output a list

dat <- read.csv( these[1], stringsAsFactors=F ) #this generates an object with the first observation.


for( i in 2:length(these) )
{

  d <- read.csv( these[i], stringsAsFactors=F )
  dat <- bind_rows( dat, d ) 

}

plot( dat$lon, dat$lat, pch=19, cex=0.5 )

plot( dat$lon, dat$lat, pch=19, cex=0.5, col=gray(0.5,0.01) )

getwd()

# write.csv( dat, "PEOPLE-GEOCODES.csv", row.names=F )

saveRDS( dat, "../Data/PEOPLE-GEOCODESv2.rds" )

setwd( ".." )  # return to main directory
```
