---
title: 'Challenge Question'
output:
  html_document:
    theme: readable
    df_print: paged
    highlight: tango
    include:
      after_body: footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=F, warning=F, eval=T)
library( dplyr )
library( pander )
```


<br>


---

<br>

# Hidden Bugs

Consider one solution to **open_goat_door()** from last week's lab. 

```{r}
open_goat_door <- function( game, a.pick )
{
   doors <- c(1,2,3)
   # if contestant selected car,
   # randomly select one of two goats 
   if( game[ a.pick ] == "car" )
   { 
     goat.doors <- doors[ game != "car" ] 
     opened.door <- sample( goat.doors, size=1 )
   }
   if( game[ a.pick ] == "goat" )
   { 
     opened.door <- doors[ game != "car" & doors != a.pick ] 
   }
   return( opened.door ) # number between 1 and 3
}
```


Note that we use an if-else structure to the code (or if-if as implemented here), to check what our initial selection is before returning the open door. 

Also note that the criteria is exactly the same no matter which scenario we are in - we can't open the door with the car, and we can't open the current pick. Why do we care if the current pick is the car? Why not create a more general solution that works for both cases? 

```{r, eval=F}
available.doors <- doors[ game != "car" & doors != a.pick ]
opened.door <- sample( available.doors, size=1 )
```

Note that we will always either have one or two doors available:

```{r}
doors <- 1:3

game <- c("goat","car","goat")
a.pick <- 1
doors[ game != "car" & doors != a.pick ]

game <- c("goat","car","goat")
a.pick <- 2
doors[ game != "car" & doors != a.pick ]

game <- c("goat","car","goat")
a.pick <- 3
doors[ game != "car" & doors != a.pick ]
```

Why does our function always work correctly when we first select the car?

```{r}
game <- c("goat","car","goat")
a.pick <- 2
available.doors <- doors[ game != "car" & doors != a.pick ]
# door 1 or 3 available

test.cases <- NULL
for( i in 1:10 )
{ 
   opened.door <- sample( available.doors, size=1 )
   test.cases <- c( test.cases, opened.door ) 
}

test.cases  # should be 1 or 3
```


But we end up with strange results when we first select a goat? 

```{r}
game <- c("goat","car","goat")
a.pick <- 1
available.doors <- doors[ game != "car" & doors != a.pick ]
# door 3 only available

test.cases <- NULL
for( i in 1:10 )
{ 
   opened.door <- sample( available.doors, size=1 )
   test.cases <- c( test.cases, opened.door ) 
}

test.cases  # should always be 3
```


This is a problem! Our function now violates all of the rules by sometimes opening Door 2 with the car behind it, and sometimes opening Door 1, our current selection. 

What is going on here???

-----------


**Can anyone figure out the cause of this very subtle but very serious bug???**

I will tell you that the logic of the above functions is correct. 

Maybe look back to the stable solution that uses the if-else structure to test initial cases. Why would that fix the code? Which scenario appears to be problematic? 

-----------


## Bug Hunters


This bug demonstrates an important concept in computer science and unit testing - that of an **edge case** - an outcome that is usually the most extreme case possible so it is not properly considered in the code, and as a result functions break when then encounter them. 

For example, if you make a payment on a bill and enter a negative number, do you receive a refund from the company? 

If a user enters text into a numeric field, will that break the code? 

Here is one example of such a case:

```{r}
x <- c( 5, 40 )
x[1]
x[2]

x[2] > x[1]  # 40 is bigger than 5

x <- c( x, "10" )  # add new number to the vector

x[2] > x[1]  # should be TRUE!
```


What happened???

Which edge case should we check for to ensure the code runs correctly here? 


<br>

-----

<br>



<br>
<br>

<style>
blockquote {
    padding: 11px 22px;
    margin: 0 0 22px;
    font-size: 18px;
    border-left: 5px solid lightgray;
}

</style>

