---
title: "Font Awesome Icons Package Test"
output:
  html_document:
    theme: readable
    df_print: paged
    highlight: tango
    css: lab-instructions.css
    include:
      after_body: footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```





## fontawesome package

### Usage: 

```{r, eval=F}
install.packages("fontawesome")
fontawesome::fa( name="github" )
fontawesome::fa( name="twitter" )
fontawesome::fa( name="home" )
```

### Output Type: 

```{r}
fa <- fontawesome::fa( name="github" )
class( fa )
str( fa )
```

```{r}
fa
```


```{r}
as.character(fa)
```

### results=asis

NO **results="asis"**. 

Prints as plain text after using paste(): 

```{r}
cat( paste0( '<a href="google.com">', fontawesome::fa( name="github" ), '</a>' ) )
```


<br>


ADD **results="asis"** 

Now it adds it to the RMD as HTML: 

```{r, results="asis"}
cat( paste0( '<a href="google.com">', fontawesome::fa( name="github" ), '</a>' ) )
```


## Stable Usage in Templates 

```{r, results="asis"}
# - get icon svg object from font awesome 
# - implicitly recast as text when added to HTML tags 
# - cat() to export, need to include {r, results="asis"}

fa <- fontawesome::fa( name="github" )
x <-  paste0( '<a href="www.google.com">  \n\n', fa, '\n\n</a>' )
cat( x )
```


<br>
<hr>
<br>




# icons package 

### Usage: 

```{r, eval=F}
remotes::install_github("mitchelloharawild/icons")
icons::download_fontawesome(version = "dev")
icons::fontawesome("twitter")
icons::fontawesome("github")
icons::fontawesome("home")
```

### Output Type

```{r}
ic <- icons::fontawesome("twitter")
class( ic )
str( ic )
```

NO **results="asis"**. 

```{r}
x <- paste0( "<a href='www.google.com'>  \n\n", as.character(ic), "\n\n</a>" )
x
```

NO **results="asis"**. 

```{r}
cat(x)
```


### results=asis 

ADD **results="asis"** 

```{r, results="asis"}
x   # results="asis"
```

ADD **results="asis"** 

```{r, results="asis"}
cat( x )  # results="asis"
```


### Add Muliple

```{r, results="asis"}
add_icons <- function()
{
  cat( as.character( icons::fontawesome("twitter") ) )
  cat( as.character( icons::fontawesome("github")  ) )
  cat( as.character( icons::fontawesome("home")    ) )
}

add_icons()
```



## Stable Usage in Templates 

The HTML links are still not functioning properly with this integration - use the fontawesome package above. 

```{r, results="asis"}
# - convert icon svg to text 
# - then add to HTML tags 
# - cat() to export, need to include {r, results="asis"}

fa <- as.character( icons::fontawesome("twitter") )
x <-  paste0( '<a href="www.google.com">  \n\n', fa, '\n\n</a>' )
cat( x )
```


<br>
<br>
<hr>
<br>
<br>
<br>
